---
title: "MPH epistasis analysis notebook"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: true
    theme: lumen
---

<style type="text/css">

body{ /* Normal  */
      font-size: 16px;
  }
td {  /* Table  */
  font-size: 16px;
}
h1.title {
  font-size: 28px;
}
h1 { /* Header 1 */
  font-size: 24px;
}
h2 { /* Header 2 */
    font-size: 22px;
}
h3 { /* Header 3 */
  font-size: 20px;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = FALSE, message = FALSE)

if (!require('tidyverse')) install.packages('tidyverse'); library('tidyverse') # for everything
if (!require('factoextra')) install.packages('factoextra'); library('factoextra') # for PCA
if (!require('ggpubr')) install.packages('ggpubr'); library('ggpubr') # for paired plots
if (!require('ggrepel')) install.packages('ggrepel'); library('ggrepel') # for labels
if (!require('ggpubr')) install.packages('ggpubr'); library('ggpubr') # for paired plots
if (!require('gridExtra')) install.packages('gridExtra'); library('gridExtra') # for arranging
if (!require('scales')) install.packages('scales'); library('scales') # for muted colors

golden = "#edae49" # Color choice
```

# First analysis

*<Insert better overview here later>*

Taking the Anc2 256 landscape and the Anc5 32 landscape files. The csvs are laid out such that each of the 8 positions is marked as '0' for wt and '1' for mutant. Then, experimental replicate rates are listed in M s-1 OD-1 of PNP formation per OD of induced cells from the lysate. I used 15000 M-1 cm-1 for the extinction coefficient for PNP which was empirically determined for pH 7.5 buffer (unlike 18000 reported in literature).

All initial rates were determined in Prism using linear fits. Replicates that showed high noise were removed, leaving data ranging from 1-10+ replicates, with the majority around n=3. Note: "initial rates" may be taken at the initial time points, or in the middle of the reaction due to "burst" or "lag" phases noted in the data. These differed among variants.

**FOR NOW burst and lag qualitative data is not annotated in the data**

## Import CSVs

```{r}
# List all CSV files in the directory
anc2 <- read_csv('anc2_rates_unclean.csv')
anc5 <- read_csv('anc5_rates_unclean.csv')
```

## Obtaining SMEs

First define the functions. Normalization to wt and log transform included here.

```{r}
library(tidyverse)

calculate_single_mutational_effects <- function(data, positions = NULL) {
  # If positions not specified, use all position columns (assuming they start with 'p')
  if (is.null(positions)) {
    positions <- names(data)[grepl("^p", names(data))]
  }
  
  # Get experimental columns
  exp_cols <- names(data)[grepl("^exp", names(data))]
  
  # Calculate stats for all genotypes
  genotype_stats <- data %>%
    select(all_of(c(positions, exp_cols))) %>%
    # Create a genotype identifier
    unite("genotype", all_of(positions), sep = "", remove = FALSE) %>%
    group_by(genotype, across(all_of(positions))) %>%
    summarise(
      across(all_of(exp_cols), ~ list(.x[!is.na(.x)])),
      .groups = "drop"
    ) %>%
    rowwise() %>%
    mutate(
      all_exp_values = list(unlist(c_across(all_of(exp_cols)))),
      mean_rate = ifelse(length(all_exp_values) > 0, mean(all_exp_values), NA),
      sd_rate = ifelse(length(all_exp_values) > 1, sd(all_exp_values), NA),
      sem_rate = ifelse(length(all_exp_values) > 1, sd(all_exp_values) / sqrt(length(all_exp_values)), NA),
      cv_rate = ifelse(!is.na(mean_rate) & mean_rate != 0 & !is.na(sd_rate), sd_rate / abs(mean_rate), NA),
      n_obs = length(all_exp_values) # This might be counting NAs
    ) %>%
    select(-all_of(exp_cols), -all_exp_values) %>%
    ungroup()
  
  # Find the all-wild-type background (all positions = 0)
  all_wt_genotype <- paste0(rep("0", length(positions)), collapse = "")
  wt_reference <- genotype_stats %>%
    filter(genotype == all_wt_genotype)
  
  if (nrow(wt_reference) != 1 || is.na(wt_reference$mean_rate)) {
    stop("Could not find valid all-wild-type reference genotype")
  }
  
  # Calculate single mutational effects
  results <- tibble()
  
  for (i in seq_along(positions)) {
    pos_name <- positions[i]
    other_positions <- positions[-i]
    
    # Get all unique backgrounds for the other positions
    if (length(other_positions) > 0) {
      backgrounds <- genotype_stats %>%
        select(all_of(other_positions)) %>%
        distinct()
      
      for (j in 1:nrow(backgrounds)) {
        background_row <- backgrounds[j, ]
        
        # Find matching genotypes
        wt_condition <- genotype_stats
        mut_condition <- genotype_stats
        
        # Filter by background
        for (pos in other_positions) {
          wt_condition <- wt_condition %>% filter(!!sym(pos) == background_row[[pos]])
          mut_condition <- mut_condition %>% filter(!!sym(pos) == background_row[[pos]])
        }
        
        # Filter by target position
        wt_condition <- wt_condition %>% filter(!!sym(pos_name) == 0)
        mut_condition <- mut_condition %>% filter(!!sym(pos_name) == 1)
        
        if (nrow(wt_condition) == 1 && nrow(mut_condition) == 1 && 
            !is.na(wt_condition$mean_rate) && !is.na(mut_condition$mean_rate) &&
            wt_condition$mean_rate > 0 && mut_condition$mean_rate > 0) {
          
          # Calculate fold change relative to all-WT background
          wt_fold_change <- wt_condition$mean_rate / wt_reference$mean_rate
          mut_fold_change <- mut_condition$mean_rate / wt_reference$mean_rate
          
          # Calculate the ratio (mutant fold change / WT fold change)
          effect_ratio <- mut_fold_change / wt_fold_change
          
          # Error propagation for ratios
          # For wt_fold_change = wt_condition / wt_reference
          cv_wt_fold <- sqrt(wt_condition$cv_rate^2 + wt_reference$cv_rate^2)
          sd_wt_fold <- wt_fold_change * cv_wt_fold
          
          # For mut_fold_change = mut_condition / wt_reference  
          cv_mut_fold <- sqrt(mut_condition$cv_rate^2 + wt_reference$cv_rate^2)
          sd_mut_fold <- mut_fold_change * cv_mut_fold
          
          # For effect_ratio = mut_fold_change / wt_fold_change
          cv_effect_ratio <- sqrt(cv_mut_fold^2 + cv_wt_fold^2)
          effect_sd <- effect_ratio * cv_effect_ratio
          
          # Calculate SEM for the ratio (approximate - assumes independent measurements)
          # This is an approximation since we're combining different sample sizes
          effective_n <- min(wt_condition$n_obs, mut_condition$n_obs, wt_reference$n_obs)
          effect_sem <- effect_sd / sqrt(effective_n)
          
          # Log-transformed values (log10 instead of log2)
          log_effect <- log10(effect_ratio)
          # For log transformation: SD_log = SD_ratio / (ratio * ln(10)) for log10
          log_effect_sd <- effect_sd / (effect_ratio * log(10))
          log_effect_sem <- effect_sem / (effect_ratio * log(10))
          
          # Create background string
          background_string <- paste0(background_row, collapse = "")
          
          # Create variant ID with X at the position being mutated
          variant_genotype <- strsplit(wt_condition$genotype, "")[[1]]
          variant_genotype[i] <- "X"
          variant_id <- paste0(variant_genotype, collapse = "")
          
          results <- bind_rows(results, tibble(
            position = pos_name,
            variant_id = variant_id,
            background = background_string,
            effect_ratio = effect_ratio,
            effect_sd = effect_sd,
            effect_sem = effect_sem,
            log10_effect = log_effect,
            log10_effect_sd = log_effect_sd,
            log10_effect_sem = log_effect_sem,
            wt_genotype = wt_condition$genotype,
            mut_genotype = mut_condition$genotype,
            wt_mean = wt_condition$mean_rate,
            mut_mean = mut_condition$mean_rate,
            wt_fold_vs_reference = wt_fold_change,
            mut_fold_vs_reference = mut_fold_change,
            wt_n = wt_condition$n_obs,
            mut_n = mut_condition$n_obs
          ))
        }
      }
    }
  }
  
  return(results)
}
```

Also useful to get epistatic effects (EEs) across all backgrounds. Normalization to wt and log transform included here.

Get it for Anc2 full

```{r}
anc2_smes <- calculate_single_mutational_effects(anc2)
```

Then Anc2 filtered for the m5 only

```{r}
anc2_partial <- anc2 %>% filter(p282 == 0, p297 == 0, p300 == 0)

anc2_partial_smes <- calculate_single_mutational_effects(anc2_partial)
```

Then Anc2 filtered for the m5+i3 only

```{r}
anc2_partial_i3 <- anc2 %>% 
  filter(p282 == 1, p297 == 1, p300 == 1) %>%
  mutate(p282 = rep(0, 32), p297 = rep(0, 32), p300 = rep(0, 32))

anc2_partial_i3_smes <- calculate_single_mutational_effects(anc2_partial_i3)
```

Then Anc5

```{r}
anc5_smes <- calculate_single_mutational_effects(anc5)
```

## Plotting

All Anc2s

```{r}
position_summary <- anc2_smes %>%
    mutate(position = factor(position, levels = c('p72', 'p193', 'p258', 'p271', 'p273', 'p282', 'p297', 'p300'))) %>%
  group_by(position) %>%
  summarise(log10_effect = mean(log10_effect))

position_summary_2 <- anc2_smes %>%
    mutate(position = factor(position, levels = c('p72', 'p193', 'p258', 'p271', 'p273', 'p282', 'p297', 'p300'))) %>%
  group_by(position) %>%
  mutate(additional_mut = str_count(background, '1')) %>%
  filter(additional_mut == 0)

anc2_smes %>%
  mutate(position = factor(position, levels = c('p72', 'p193', 'p258', 'p271', 'p273', 'p282', 'p297', 'p300'))) %>%
  ggplot(aes(x = position, y = log10_effect)) +
  geom_hline(yintercept = 0, lty = 2) +
  #geom_hline(yintercept = log10(3.4), lty = 2) +
  #geom_hline(yintercept = log10(1/3.4), lty = 2) +
  geom_jitter(width = 0.2, color = "lightblue") +
  geom_point(data = position_summary, aes(y = log10_effect), 
             color = "darkred", size = 2) +
  geom_point(data = position_summary_2, aes(y = log10_effect), 
             color = "black", size = 2) +
  labs(x = "Position",
       y = "log10 fold-change") +
  ylim(c(-4, 4)) +
  theme_classic()
```

Anc2 partial

```{r}
anc2_partial_smes %>%
  mutate(position = factor(position, levels = c('p72', 'p193', 'p258', 'p271', 'p273', 'p282', 'p297', 'p300'))) %>%
  ggplot(aes(x = position, y = log10_effect)) +
  geom_hline(yintercept = 0) +
  geom_hline(yintercept = log10(3.4), lty = 2) +
  geom_hline(yintercept = log10(1/3.4), lty = 2) +
  geom_jitter(width = 0.2, color = "lightblue") +
  #geom_point(data = position_summary, aes(y = mean_effect), 
  #           color = "darkred", size = 2) +
  #geom_errorbar(aes(ymin = effect - effect_sd,
  #                  ymax = effect + effect_sd),
  #              color = "red", width = 0.2) +
  labs(x = "Position",
       y = "log10 fold-change") +
  ylim(c(-3, 3)) +
  theme_classic()
```

Anc5

```{r}
anc5_smes %>%
  mutate(position = factor(position, levels = c('p72', 'p193', 'p258', 'p271', 'p273', 'p282', 'p297', 'p300'))) %>%
  ggplot(aes(x = position, y = log10_effect)) +
  geom_hline(yintercept = 0) +
  geom_hline(yintercept = log10(3.4), lty = 2) +
  geom_hline(yintercept = log10(1/3.4), lty = 2) +
  geom_jitter(width = 0.2, color = "lightblue") +
  #geom_point(data = position_summary, aes(y = mean_effect), 
  #           color = "darkred", size = 2) +
  #geom_errorbar(aes(ymin = effect - effect_sd,
  #                  ymax = effect + effect_sd),
  #              color = "red", width = 0.2) +
  labs(x = "Position",
       y = "log10 fold-change") +
  ylim(c(-3, 3)) +
  theme_classic()
```

Anc2+i3

```{r}
anc2_partial_i3_smes %>%
  mutate(position = factor(position, levels = c('p72', 'p193', 'p258', 'p271', 'p273', 'p282', 'p297', 'p300'))) %>%
  ggplot(aes(x = position, y = log10_effect)) +
  geom_hline(yintercept = 0) +
  geom_hline(yintercept = log10(3.4), lty = 2) +
  geom_hline(yintercept = log10(1/3.4), lty = 2) +
  geom_jitter(width = 0.2, color = "lightblue") +
  #geom_point(data = position_summary, aes(y = mean_effect), 
  #           color = "darkred", size = 2) +
  #geom_errorbar(aes(ymin = effect - effect_sd,
  #                  ymax = effect + effect_sd),
  #              color = "red", width = 0.2) +
  labs(x = "Position",
       y = "log10 fold-change") +
  ylim(c(-3, 3)) +
  theme_classic()
```

Combined Anc2 and Anc5

```{r}
# Combine the datasets with ancestry labels
anc2_labeled <- anc2_partial_smes %>% mutate(ancestry = "anc2")
anc5_labeled <- anc5_smes %>% mutate(ancestry = "anc5")

combined_effects <- bind_rows(anc2_labeled, anc5_labeled)
```

With lines connecting

```{r}
# Create manual x-position adjustments
dodge_width <- 0.3
combined_effects <- combined_effects %>%
  mutate(
    position = factor(position, levels = c('p72', 'p193', 'p258', 'p271', 'p273')),
    position_numeric = as.numeric(position),
    x_position = case_when(
      ancestry == "anc2" ~ position_numeric - dodge_width/2,
      ancestry == "anc5" ~ position_numeric + dodge_width/2
    )
  )

# Create the plot with manual positioning
p <- combined_effects %>%
  mutate(ancestry = case_when(ancestry == "anc2" ~ "AncDHCH2",
                              ancestry == "anc5" ~ "AncDHCH5")) %>%
  ggplot(aes(x = x_position, y = log10_effect, color = ancestry)) +
  # Add horizontal line at y = 0 (no effect) and thresholds (currently hardcoded)
  geom_hline(yintercept = 0, alpha = 0.5) +
  geom_hline(yintercept = log10(2), linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = log10(1/2), linetype = "dashed", alpha = 0.5) +
  # Add connecting lines between anc2 and anc5 for the same variant_id
  geom_line(aes(group = variant_id), 
            color = "black", linewidth = 0.3) +
  # Add points
  geom_point(size = 1.5, alpha = 0.7) +
  # Add error bars
  geom_errorbar(aes(ymin = log10_effect - log10_effect_sem, 
                    ymax = log10_effect + log10_effect_sem),
                width = 0.05) +
  # Customize colors
  scale_color_manual(values = c("AncDHCH2" = "#E31A1C", "AncDHCH5" = "#1F78B4")) +
  # Set x-axis to show position names
  scale_x_continuous(
    breaks = 1:length(unique(combined_effects$position)),
    labels = unique(combined_effects$position)[order(unique(combined_effects$position))]
  ) +
  # Labels and theme
  labs(
    x = "Position",
    y = expression('log'[10]*' SME (single mutational effect)'),
    color = "Background",
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 8),
    axis.title = element_text(size = 8),
    panel.grid.minor = element_blank()
  )

p

# With legend
ggsave("anc2_vs_anc5_smes_leg.svg", p, width = 180/2, height = 247/4.2, dpi = 600, units = "mm")

# Without legend
ggsave("anc2_vs_anc5_smes.svg", p+theme(legend.position = "none"), width = 180/2, height = 247/4.2, dpi = 600, units = "mm")

write_csv(combined_effects %>%
  mutate(ancestry = case_when(ancestry == "anc2" ~ "AncDHCH2",
                              ancestry == "anc5" ~ "AncDHCH5")), "smes_anc2_anc5.csv")
```

Combined Anc2+i3 and Anc5

```{r}
# Combine the datasets with ancestry labels
anc2_i3_labeled <- anc2_partial_i3_smes %>% mutate(ancestry = "anc2+i3")

combined_effects_2 <- bind_rows(anc2_i3_labeled, anc5_labeled)
```

With lines connecting

```{r}
# Create manual x-position adjustments
combined_effects_2 <- combined_effects_2 %>%
  mutate(
    position = factor(position, levels = c('p72', 'p193', 'p258', 'p271', 'p273')),
    position_numeric = as.numeric(position),
    x_position = case_when(
      ancestry == "anc2+i3" ~ position_numeric - dodge_width/2,
      ancestry == "anc5" ~ position_numeric + dodge_width/2
    )
  )

# Create the plot with manual positioning
p_2 <- combined_effects_2 %>%
  ggplot(aes(x = x_position, y = log10_effect, color = ancestry)) +
  # Add horizontal line at y = 0 (no effect) and thresholds (currently hardcoded)
  geom_hline(yintercept = 0, alpha = 0.5) +
  geom_hline(yintercept = log10(2), linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = log10(1/2), linetype = "dashed", alpha = 0.5) +
  # Add connecting lines between anc2 and anc5 for the same variant_id
  geom_line(aes(group = variant_id), 
            color = "black", linewidth = 0.3) +
  # Add points
  geom_point(size = 2, alpha = 0.7) +
  # Add error bars
  geom_errorbar(aes(ymin = log10_effect - log10_effect_sem, 
                    ymax = log10_effect + log10_effect_sem),
                width = 0.05) +
  # Customize colors
  scale_color_manual(values = c("anc2+i3" = "#E31A1C", "anc5" = "#1F78B4")) +
  # Set x-axis to show position names
  scale_x_continuous(
    breaks = 1:length(unique(combined_effects$position)),
    labels = unique(combined_effects$position)[order(unique(combined_effects$position))]
  ) +
  # Labels and theme
  labs(
    x = "Position",
    y = expression('log'[10]*' SME (single mutational effect)'),
    color = "Ancestry",
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )

p_2

ggsave("anc2_i3_vs_anc5_smes.svg", p_2, width = 180, height = 247/3, dpi = 300, units = "mm")
```

Combined Anc2 and Anc2+i3

```{r}
# Combine the datasets with ancestry labels
combined_effects_3 <- bind_rows(anc2_labeled, anc2_i3_labeled)
```

With lines connecting

```{r}
# Create manual x-position adjustments
combined_effects_3 <- combined_effects_3 %>%
  mutate(
    position = factor(position, levels = c('p72', 'p193', 'p258', 'p271', 'p273')),
    position_numeric = as.numeric(position),
    x_position = case_when(
      ancestry == "anc2" ~ position_numeric - dodge_width/2,
      ancestry == "anc2+i3" ~ position_numeric + dodge_width/2
    )
  )

# Create the plot with manual positioning
p_3 <- combined_effects_3 %>%
  ggplot(aes(x = x_position, y = log10_effect, color = ancestry)) +
  # Add horizontal line at y = 0 (no effect) and thresholds (currently hardcoded)
  geom_hline(yintercept = 0, alpha = 0.5) +
  geom_hline(yintercept = log10(2), linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = log10(1/2), linetype = "dashed", alpha = 0.5) +
  # Add connecting lines between anc2 and anc5 for the same variant_id
  geom_line(aes(group = variant_id), 
            color = "black", linewidth = 0.3) +
  # Add points
  geom_point(size = 2, alpha = 0.7) +
  # Add error bars
  geom_errorbar(aes(ymin = log10_effect - log10_effect_sem, 
                    ymax = log10_effect + log10_effect_sem),
                width = 0.05) +
  # Customize colors
  scale_color_manual(values = c("anc2" = "#E31A1C", "anc2+i3" = "#1F78B4")) +
  # Set x-axis to show position names
  scale_x_continuous(
    breaks = 1:length(unique(combined_effects$position)),
    labels = unique(combined_effects$position)[order(unique(combined_effects$position))]
  ) +
  # Labels and theme
  labs(
    x = "Position",
    y = expression('log'[10]*' SME (single mutational effect)'),
    color = "Ancestry",
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )

p_3

#ggsave("anc2_vs_anc2_i3s_smes.svg", p_3, width = 180, height = 247/3, dpi = 300, units = "mm")
```

Also just the 282/300 combo with AncDHCH5

```{r}
anc2_partial_two_i3 <- anc2 %>% 
  filter(p282 == 1, p297 == 0, p300 == 1) %>%
  mutate(p282 = rep(0, 32), p297 = rep(0, 32), p300 = rep(0, 32))

anc2_partial_two_i3 <- calculate_single_mutational_effects(anc2_partial_two_i3)

anc2_two_i3_labeled <- anc2_partial_two_i3 %>% mutate(ancestry = "anc2+282/300")

combined_effects_4 <- bind_rows(anc2_two_i3_labeled, anc5_labeled)
```

With lines connecting

```{r}
# Create manual x-position adjustments
dodge_width <- 0.3
combined_effects_4 <- combined_effects_4 %>%
  mutate(
    position = factor(position, levels = c('p72', 'p193', 'p258', 'p271', 'p273')),
    position_numeric = as.numeric(position),
    x_position = case_when(
      ancestry == "anc2+282/300" ~ position_numeric - dodge_width/2,
      ancestry == "anc5" ~ position_numeric + dodge_width/2
    )
  )

# Create the plot with manual positioning
p <- combined_effects_4 %>%
  mutate(ancestry = case_when(ancestry == "anc2+282/300" ~ "AncDHCH2+282|300",
                              ancestry == "anc5" ~ "AncDHCH5")) %>%
  ggplot(aes(x = x_position, y = log10_effect, color = ancestry)) +
  # Add horizontal line at y = 0 (no effect) and thresholds (currently hardcoded)
  geom_hline(yintercept = 0, alpha = 0.5) +
  geom_hline(yintercept = log10(2), linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = log10(1/2), linetype = "dashed", alpha = 0.5) +
  # Add connecting lines between anc2 and anc5 for the same variant_id
  geom_line(aes(group = variant_id), 
            color = "black", linewidth = 0.3) +
  # Add points
  geom_point(size = 1.5, alpha = 0.7) +
  # Add error bars
  geom_errorbar(aes(ymin = log10_effect - log10_effect_sem, 
                    ymax = log10_effect + log10_effect_sem),
                width = 0.05) +
  # Customize colors
  scale_color_manual(values = c("AncDHCH2+282|300" = "#E31A1C", "AncDHCH5" = "#1F78B4")) +
  # Set x-axis to show position names
  scale_x_continuous(
    breaks = 1:length(unique(combined_effects$position)),
    labels = unique(combined_effects$position)[order(unique(combined_effects$position))]
  ) +
  # Labels and theme
  labs(
    x = "Position",
    y = expression('log'[10]*' SME (single mutational effect)'),
    color = "Background",
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 8),
    axis.title = element_text(size = 8),
    panel.grid.minor = element_blank()
  )

p

# With legend
ggsave("anc2_i2_vs_anc5_smes_leg.svg", p, width = 180/2.5, height = 247/4.2, dpi = 600, units = "mm")

# Without legend
ggsave("anc2_i2_vs_anc5_smes.svg", p+theme(legend.position = "none"), width = 180/2.5, height = 247/4.2, dpi = 600, units = "mm")

write_csv(combined_effects_4 %>%
  mutate(ancestry = case_when(ancestry == "anc2+282/300" ~ "AncDHCH2+282|300",
                              ancestry == "anc5" ~ "AncDHCH5")), "smes_anc2_282_300_anc5.csv")
```

# Multisub inspired analysis

Using some of the code from the multisubstrate analysis Rmd for PTE

## Clean up

Also adding Anc5 for later

```{r}
processed_anc2 <- anc2 %>%
  pivot_longer(cols = exp1:exp23, names_to = 'replicate', values_to = 'activity', values_drop_na = TRUE) %>%
  # "L∆HFIALA" is wt while "RSLLTPYV" is mutant
  mutate(g72 = if_else(p72 == 0, "L", "R"),
       g193 = if_else(p193 == 0, "∆", "S"),
       g258 = if_else(p258 == 0, "H", "L"),
       g271 = if_else(p271 == 0, "F", "L"),
       g273 = if_else(p273 == 0, "I", "T"),
       g282 = if_else(p282 == 0, "A", "P"),
       g297 = if_else(p297 == 0, "L", "Y"),
       g300 = if_else(p300 == 0, "A", "V")) %>%
  unite("reference", g72:g300, na.rm = TRUE, remove = TRUE, sep = "")

processed_anc5 <- anc5 %>%
  pivot_longer(cols = exp1:exp6, names_to = 'replicate', values_to = 'activity', values_drop_na = TRUE) %>%
  # "L∆HFIALA" is wt while "RSLLTPYV" is mutant
  mutate(g72 = if_else(p72 == 0, "L", "R"),
       g193 = if_else(p193 == 0, "∆", "S"),
       g258 = if_else(p258 == 0, "H", "L"),
       g271 = if_else(p271 == 0, "F", "L"),
       g273 = if_else(p273 == 0, "I", "T"),
       g282 = if_else(p282 == 1, "A", "P"),
       g297 = if_else(p297 == 1, "L", "Y"),
       g300 = if_else(p300 == 1, "A", "V")) %>%
  unite("reference", g72:g300, na.rm = TRUE, remove = TRUE, sep = "")
```

```{r}
mae_mean <- function(x) {
  abs_devs = c()
  for(i in 1:length(x)) {
    abs_devs = c(abs(x - mean(x)), abs_devs)
  }
  return(mean(abs_devs))
}

mae_median <- function(x) {
  abs_devs = c()
  for(i in 1:length(x)) {
    abs_devs = c(abs(x - median(x)), abs_devs)
  }
  return(median(abs_devs))
}


processed_mean_anc2 <- processed_anc2 %>%
  group_by(p72, p193, p258, p271, p273, p282, p297, p300, reference) %>%
  summarise(mean_activity = mean(activity),
            sd_activity = sd(activity),
            mae_mean_activity = mae_mean(activity),
            mae_median_activity = mae_median(activity)) %>%
  mutate(cv = sd_activity/mean_activity*100,
         mean_fold_error = (mae_mean_activity/mean_activity) + 1,
         median_absolute_deviation = mae_median_activity/mean_activity)

processed_mean_anc5 <- processed_anc5 %>%
  group_by(p72, p193, p258, p271, p273, p282, p297, p300, reference) %>%
  summarise(mean_activity = mean(activity),
            sd_activity = sd(activity),
            mae_mean_activity = mae_mean(activity),
            mae_median_activity = mae_median(activity)) %>%
  mutate(cv = sd_activity/mean_activity*100,
         mean_fold_error = (mae_mean_activity/mean_activity) + 1,
         median_absolute_deviation = mae_median_activity/mean_activity)

```

We can then look at the distribution of coefficient of variation for each variant

```{r}
cv_fig <- processed_mean_anc2 %>%
  ggplot(aes(x = cv)) +
  geom_histogram(fill = golden) +
  labs(x = "Coefficient of Variation (%)", y = "Genotypes") +
  theme_classic()

cv_fig
```

And mean average error

```{r}
mae_fig <- processed_mean_anc2 %>%
  ggplot(aes(x = mean_fold_error)) +
  geom_histogram(fill = golden) +
  geom_vline(xintercept = 1.5, lty = 2) +
  geom_vline(xintercept = 2, lty = 2, col = "darkred") +
  labs(x = "Mean fold error (MAE)", y = "Genotypes") +
  theme_classic()

mae_fig
```

## Creating a normalized dataframe

### Ensuring log-transformed data is normally distributed

As the title says. To do this, it is worth collecting all the data, subtracting by the mean (to set the mean equal to 0) for each reference/substrate pair, then pooling all the data to verify that it is normally distributed (i.e. the errors on the log scale are normally distributed)

```{r}
norm_data <- processed_anc2 %>%
  mutate(norm_activity = activity/mean(activity[reference == "L∆HFIALA"]), # Normalize by the mean activity of the wt sequence
         norm_activity = log10(norm_activity)) %>% # Log-transform
  ungroup() %>%
  group_by(reference) %>%
  mutate(norm_value = norm_activity - mean(norm_activity)) %>% # Set mean of each measurement to 0 
  ungroup() 

norm_fig <- norm_data %>%
  ggplot(aes(x = norm_value)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 0.01, color = "black", fill = golden, lwd = 0.2) +
  stat_function(fun = dnorm, args = list(mean = mean(norm_data$norm_value), sd = sd(norm_data$norm_value)), colour = "black") +
  theme_classic()

norm_fig

##ggsave("norm_fig.svg", plot = norm_fig, width = 180, height = 247/3, dpi = 300, units = "mm")
```

And the qqplot

```{r}
qqnorm(norm_data$norm_value, pch = 1, frame = FALSE)
qqline(norm_data$norm_value, col = "steelblue", lwd = 2)
```

### Creating a data frame with statistics

It may be useful to have a data frame that describes all of the general statistics. Since we've tested for normality, it is fair to get a mean and SD directly from the log-transformed data. **Note:** when you use these values to convert back to a fold-change scale, you will obtain a geometric mean of the fold-change. This is probably fair as "fold-change errors" are likely log-normal, this GM and GMSD are better metrics for those.

```{r normalizing-no-norm-mae-hist}
normalized_bulky_mean_data <- processed_anc2 %>%
  mutate(norm_activity = log10(activity), # Log-transform
         norm_activity = norm_activity - mean(norm_activity[reference == "L∆HFIALA"])) %>% # Normalize by the mean activity of the wt sequence
  ungroup() %>%
  select(-c("replicate", "activity")) %>%
  group_by(p72, p193, p258, p271, p273, p282, p297, p300, reference) %>%
  # Error propagation - http://surl.li/qrptg
  summarize(mean_value = mean(norm_activity),
         sd_value = sd(norm_activity),
         total_entries = n()) %>%
  mutate(se_value = sd_value / sqrt(total_entries) ) %>% ## THIS WAS NOT SQUARE ROOT PREVIOUSLY, FIXED
  ungroup()

normalized_bulky_mean_data_anc5 <- processed_anc5 %>%
  mutate(norm_activity = log10(activity), # Log-transform
         norm_activity = norm_activity - mean(norm_activity[reference == "L∆HFIALA"])) %>% # Normalize by the mean activity of the wt sequence
  ungroup() %>%
  select(-c("replicate", "activity")) %>%
  group_by(p72, p193, p258, p271, p273, p282, p297, p300, reference) %>%
  # Error propagation - http://surl.li/qrptg
  summarize(mean_value = mean(norm_activity),
         sd_value = sd(norm_activity),
         total_entries = n()) %>%
  mutate(se_value = sd_value / sqrt(total_entries)) %>%
  ungroup()
```

## Epistasis tests

### Generating SMEs and EEs

```{r}
## Helper functions 

error_prop <- function(x){
  return(sqrt(sum(x**2)))
}

harm_mean <- function(x){
  return(length(x)/sum(1/x))
}

alternate_sub_add <- function(x){
  # Takes the last term and alternates between subtracting and adding the next term to the left
  y = x[length(x)]
  counter = T
  if(length(x) > 1) {
    for(i in (length(x)-1):1){
      if(counter){
        y = y - x[i]
        counter = F
      } else{
        y = y + x[i]
        counter = T
      }
    }
  }
  return(y)
}

# Generate all possible combinations of numbers 1 to 8
all_combinations <- lapply(1:8, function(x) combn(1:8, x, simplify = FALSE))

# Combine all combinations into a single list
combined_combinations <- unlist(all_combinations, recursive = FALSE)

# Add the combination with all numbers
combined_combinations <- c(combined_combinations, list(1:8))

###

ees = data.frame()

for(combo in 1:length(combined_combinations)) {
  if(combo != length(combined_combinations)) {
    temp_small_data_ready <- normalized_bulky_mean_data %>%
    rowwise() %>%
    mutate(muts = sum(p72, p193, p258, p271, p273, p282, p297, p300)) %>%
    ungroup() %>%
    filter_at(vars(combined_combinations[[combo]]), all_vars(. != 1)) %>%
    group_by(muts)
  } else {
    temp_small_data_ready <- normalized_bulky_mean_data %>%
    rowwise() %>%
    mutate(muts = sum(p72, p193, p258, p271, p273, p282, p297, p300)) %>%
    ungroup() %>%
    group_by(muts)
  }

  
  cur_harm_mean <- temp_small_data_ready %>% pull(total_entries) %>% harm_mean() # Get harmonic mean
  
  # Calculate epistasis with error propagation
  next_term <- temp_small_data_ready %>%
    summarize(new_val = sum(mean_value), ## Should I check this process?
              new_se = error_prop(se_value)) %>%
    ungroup() %>%
    summarize(new_val = alternate_sub_add(new_val),
              new_se = error_prop(new_se)) %>%
    mutate(new_n = cur_harm_mean)
  
  ref_geno = "00000000"
  current = unique(which(!(seq(1, 8, 1) %in% combined_combinations[[combo]])))
  if(length(current) > 0){
    for(number in 1:length(current)){
      substr(ref_geno, current[number], current[number]) <- 'x'
    }
  } else {
    ref_geno = "xxxxxxxx"
  }
  
  ees = rbind(ees, next_term %>% mutate(geno = ref_geno))
}

ees <- ees %>% distinct() # Remove duplicates

# Remove artifactual second to last order row
ees <- ees %>% 
  slice(- (n() - 1))
```

### Significance testing of EEs

Next comes significance testing. Thanks to Dan for the code to perform the adjusted p-value.

```{r ee-sig-test}
## SOME NA VALUES? WHY? Probably because I am missing replicates for certain genotypes
test_ees <- ees %>%
  mutate(t = (new_val)/new_se,
         p = 2*pt(abs(t),df=new_n-1,lower=F)) %>%
  mutate(padj = p.adjust(p,method = 'fdr'),n=nrow(.),
         muts = str_count(geno, 'x'),
         sig = padj < 0.01)
```

Finally we can visualize this in volcano plots to look at EEs (x-axis) vs significance (y-axis):

```{r}
volcano_fig <- test_ees %>%
  filter(muts > 1) %>%
  mutate(muts = factor(muts)) %>%
  ggplot(aes(x = new_val, y = -1*log10(padj), color = muts)) +
  geom_hline(yintercept = -log10(0.01), lty = 2) +
  geom_vline(xintercept = 0, lty = 2) +
  geom_point() +
  #xlim(-2.5, 2.5) +
  labs(x = "log10 Epistatic Coefficient", 
       y = expression("-log"[10]*" p-value"),
       color = "Mutations") +
  theme_classic()

volcano_fig

##ggsave("volcano_fig.svg", plot = volcano_fig, width = 180, height = 247/2, dpi = 300, units = "mm")

```

First it is worth verifying single mutational effects look the same as previous analysis

```{r}
sme_sub_fig <- test_ees %>%
  filter(muts == 1) %>%
  mutate(muts = as.factor(muts)) %>%
  ggplot(aes(x = muts, y = new_val, col = sig)) +
  geom_point() +
  geom_errorbar(aes(ymin=new_val-new_se, ymax=new_val+new_se), 
                width=.05, 
                position=position_dodge(.9)) +
  geom_hline(yintercept = 0, lty = 2) +
  facet_wrap(~geno) +
  theme_classic() +
  labs(x = "Mutations", y = "Epistasis") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

sme_sub_fig

##ggsave("sme_sub_fig.svg", plot = sme_sub_fig, width = 180, height = 247/2, dpi = 300, units = "mm")
```

And second

```{r coef-by-sub-2nd}
ee2_sub_fig <- test_ees %>%
  filter(muts == 2) %>%
  mutate(muts = as.factor(muts)) %>%
  ggplot(aes(x = muts, y = new_val, col = sig)) +
  geom_point() +
  geom_errorbar(aes(ymin=new_val-new_se, ymax=new_val+new_se), 
                width=.05, 
                position=position_dodge(.9)) +
  geom_hline(yintercept = 0, lty = 2) +
  facet_wrap(~geno) +
  theme_classic() +
  labs(x = "Mutations", y = "Epistasis") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ee2_sub_fig

##ggsave("ee2_sub_figb_fig.svg", plot = ee2_sub_fig, width = 180, height = 247/2, dpi = 300, units = "mm")
```

And third

```{r coef-by-sub-3rd}
ee2_sub_fig <- test_ees %>%
  filter(muts == 3) %>%
  mutate(muts = as.factor(muts)) %>%
  ggplot(aes(x = muts, y = new_val, col = sig)) +
  geom_point() +
  geom_errorbar(aes(ymin=new_val-new_se, ymax=new_val+new_se), 
                width=.05, 
                position=position_dodge(.9)) +
  geom_hline(yintercept = 0, lty = 2) +
  facet_wrap(~geno) +
  theme_classic() +
  labs(x = "Mutations", y = "Epistasis") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ee2_sub_fig

##ggsave("ee3_sub_figb_fig.svg", plot = ee3_sub_fig, width = 180, height = 247/2, dpi = 300, units = "mm")
```

## i3 on m5 modulation

Here I run significance testing on subsets of the data, starting with 282 mutated

```{r}

get_ees_subset <- function(current_data, m282, m297, m300, type) {
  normalized_bulky_mean_data_temp <- current_data %>%
  filter(p282 == m282, p297 == m297, p300 == m300)

  wt_ref <- (normalized_bulky_mean_data_temp %>% filter(p72 == 0, p193 == 0, p258 == 0, p271 == 0, p273 == 0))$reference[1]
  
  normalized_bulky_mean_data_temp <- normalized_bulky_mean_data_temp %>%
    mutate(norm_activity = log10(activity), # Log-transform
           norm_activity = norm_activity - mean(norm_activity[reference == wt_ref])) %>% # Normalize by the mean activity of the wt sequence
    ungroup() %>%
    select(-c("replicate", "activity")) %>%
    group_by(p72, p193, p258, p271, p273, reference) %>%
    # Error propagation - http://surl.li/qrptg
    summarize(mean_value = mean(norm_activity),
           sd_value = sd(norm_activity),
           total_entries = n()) %>%
    mutate(se_value = sd_value / sqrt(total_entries)) %>%
    ungroup()
  
  all_combinations <- lapply(1:5, function(x) combn(1:5, x, simplify = FALSE))
  
  # Combine all combinations into a single list
  combined_combinations <- unlist(all_combinations, recursive = FALSE)
  
  # Add the combination with all numbers
  combined_combinations <- c(combined_combinations, list(1:5))
  
  ###
  
  ees = data.frame()
  
  for(combo in 1:length(combined_combinations)) {
    if(combo != length(combined_combinations)) {
      temp_small_data_ready <- normalized_bulky_mean_data_temp %>%
      rowwise() %>%
      mutate(muts = sum(p72, p193, p258, p271, p273)) %>%
      ungroup() %>%
      filter_at(vars(combined_combinations[[combo]]), all_vars(. != 1)) %>%
      group_by(muts)
    } else {
      temp_small_data_ready <- normalized_bulky_mean_data_temp %>%
      rowwise() %>%
      mutate(muts = sum(p72, p193, p258, p271, p273)) %>%
      ungroup() %>%
      group_by(muts)
    }

    cur_harm_mean <- temp_small_data_ready %>% pull(total_entries) %>% harm_mean() # Get harmonic mean
    
    # Calculate epistasis with error propagation
    next_term <- temp_small_data_ready %>%
      summarize(new_val = sum(mean_value),
                new_se = error_prop(se_value)) %>%
      ungroup() %>%
      summarize(new_val = alternate_sub_add(new_val),
                new_se = error_prop(new_se)) %>%
      mutate(new_n = cur_harm_mean)
    
    ref_geno = "00000"
    current = unique(which(!(seq(1, 5, 1) %in% combined_combinations[[combo]])))
    if(length(current) > 0){
      for(number in 1:length(current)){
        substr(ref_geno, current[number], current[number]) <- 'x'
      }
    } else {
      ref_geno = "xxxxx"
    }
    
    ees = rbind(ees, next_term %>% mutate(geno = ref_geno))
  }
  
  ees <- ees %>% distinct() # Remove duplicates
  
  # Remove artifactual second to last order row
  ees <- ees %>% 
    slice(- (n() - 1))
  
  test_ees <- ees %>%
  mutate(t = (new_val)/new_se,
         p= 2*pt(abs(t),df=new_n-1,lower=F)) %>%
  mutate(padj = p.adjust(p,method = 'fdr'),n=nrow(.),
         muts = str_count(geno, 'x'),
         sig = padj < 0.05,
         type = type)
  
  return(test_ees)
}

ees_wt <- get_ees_subset(processed_anc2, 0, 0, 0, 'wt')
ees_282 <- get_ees_subset(processed_anc2, 1, 0, 0, '282')
ees_297 <- get_ees_subset(processed_anc2, 0, 1, 0, '297')
ees_300 <- get_ees_subset(processed_anc2, 0, 0, 1, '300')
ees_282_297 <- get_ees_subset(processed_anc2, 1, 1, 0, '282|297')
ees_282_300 <- get_ees_subset(processed_anc2, 1, 0, 1, '282|300')
ees_297_300 <- get_ees_subset(processed_anc2, 0, 1, 1, '297|300')
ees_282_297_300 <- get_ees_subset(processed_anc2, 1, 1, 1, '282|297|300')
ees_anc5 <- get_ees_subset(processed_anc5, 0, 0, 0, 'anc5')
```

Now I need to make a heat map where each column is each ees_set. First I need to move the last entry (xxxxx) to be the first entry, then flip the entire dataset, then take the new_val and sig

```{r}
ees_wt_moved <- ees_wt %>% 
  slice(n(), 1:(n()-1)) %>%
  arrange(desc(row_number())) %>%
  select(c(new_val, geno, sig, type))

ees_282_moved <- ees_282 %>% 
  slice(n(), 1:(n()-1)) %>%
  arrange(desc(row_number())) %>%
  select(c(new_val, geno, sig, type))

ees_297_moved <- ees_297 %>% 
  slice(n(), 1:(n()-1)) %>%
  arrange(desc(row_number())) %>%
  select(c(new_val, geno, sig, type))

ees_300_moved <- ees_300 %>% 
  slice(n(), 1:(n()-1)) %>%
  arrange(desc(row_number())) %>%
  select(c(new_val, geno, sig, type))

ees_282_297_moved <- ees_282_297 %>% 
  slice(n(), 1:(n()-1)) %>%
  arrange(desc(row_number())) %>%
  select(c(new_val, geno, sig, type))

ees_282_300_moved <- ees_282_300 %>% 
  slice(n(), 1:(n()-1)) %>%
  arrange(desc(row_number())) %>%
  select(c(new_val, geno, sig, type))

ees_297_300_moved <- ees_297_300 %>% 
  slice(n(), 1:(n()-1)) %>%
  arrange(desc(row_number())) %>%
  select(c(new_val, geno, sig, type))

ees_282_297_300_moved <- ees_282_297_300 %>% 
  slice(n(), 1:(n()-1)) %>%
  arrange(desc(row_number())) %>%
  select(c(new_val, geno, sig, type))

ees_anc5_moved <- ees_anc5 %>% 
  slice(n(), 1:(n()-1)) %>%
  arrange(desc(row_number())) %>%
  select(c(new_val, geno, sig, type))
```

Then visualize

```{r}
ees_combined <- bind_rows(list(ees_wt_moved, 
                               ees_282_moved, 
                               ees_297_moved, 
                               ees_300_moved, 
                               ees_282_297_moved,
                               ees_282_300_moved,
                               ees_297_300_moved,
                               ees_282_297_300_moved,
                               ees_anc5))
                      
# 2. Plotting
ees_combined %>%
  mutate(sig = case_when(is.na(sig) ~ FALSE,
                         TRUE ~ sig),
         geno = fct_inorder(geno),
         type = fct_inorder(type)) %>%
  ggplot(aes(x = type, y = geno, fill = new_val)) +
  # Use 'color' for the border and 'linewidth' to make it stand out
  geom_tile(aes(color = sig), width = 0.9, height = 0.9, linewidth = 0.3) +
  # Blue-to-Red gradient
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", 
                       midpoint = mean(ees_combined$new_val, na.rm = TRUE)) +
  # Define border colors: e.g., black for TRUE, transparent or light grey for FALSE
  scale_color_manual(values = c("TRUE" = "black", "FALSE" = "transparent")) +
  theme_minimal() +
  theme(
    # Tilt x-axis labels 45 degrees
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    panel.grid = element_blank()
  ) +
  labs(fill = "Value", color = "Significant") +
  coord_fixed() # Keeps tiles square
```

Then different view

```{r}
ees_combined_for_plot <- ees_combined %>%
  mutate(type = case_when(type == "wt" ~ "AncDHCH2",
                          type == "282" ~ "+282",
                          type == "297" ~ "+297",
                          type == "300" ~ "+300",
                          type == "282|297" ~ "+282|297",
                          type == "282|300" ~ "+282|300",
                          type == "297|300" ~ "+297|300",
                          type == "282|297|300" ~ "+282|297|300",
                          type == "anc5" ~ "AncDHCH5")) %>%
  mutate(
    sig = ifelse(is.na(sig), FALSE, sig),
    geno = fct_reorder(geno, muts, .na_rm = TRUE), # THIS MIGHT BE WRONG, CHECK IF NEEDED
    type = fct_rev(fct_inorder(type)) # Reversed so AncDHCH2 is on top if desired
  )

tile_2 <- ees_combined_for_plot %>%
  ggplot(aes(x = geno, y = type, fill = new_val)) +
  # Black border for significance, light grey for others to match the screenshot "grid" look
  geom_tile(aes(color = sig), width = 0.9, height = 0.9, linewidth = 0.3) +
  # Moving X axis to the top
  scale_x_discrete(position = "top") +
  scale_fill_gradient2(
    low = "#1F78B4", mid = "white", high = "#E31A1C",
    midpoint = mean(ees_combined$new_val, na.rm = TRUE)
  ) +
  scale_color_manual(values = c("TRUE" = "black", "FALSE" = "grey90")) +
  theme_minimal() +
  theme(
    # 45 degree tilt, adjusted for top position
    axis.text.x.top = element_text(angle = 45, hjust = 0, vjust = 0, size = 8),
    axis.title.y = element_blank(),
    axis.text.y = element_text(size = 8),
    panel.grid = element_blank(),
    legend.position = "right",
    legend.text = element_text(size = 8)
  ) +
  labs(x = "Genotype", 
       fill = "Epistatic effect (EE)") +
  coord_fixed()

tile_2

# Clean figure
ggsave("fig_4_e.svg", plot = tile_2 + theme(legend.position = "none"), width = 180/1.2, height = 247/4, dpi = 600, units = "mm")
# With legend for reference
ggsave("fig_4_e_legend.svg", plot = tile_2, width = 180/1.2, height = 247/4, dpi = 600, units = "mm")

write_csv(ees_combined_for_plot, "all_smes_and_ees.csv")

```

It appears that the i3s are doing a few subtle things:
1) Disrupting SME of 193, 258, and 271... This becomes primarily apparent in 282|300 and 282|297|300
2) They are amplifying the negative pairwise EEs of the m5 (x000x, 00x0x)
3) Less significant positive epistasis in the 3rd order

Note these are patterns of epistasis, not phenotypes. These are normalized to their respective WT and the i3s show subtle changes to "wt" activity.

## Heat map for landscapes

```{r}
combo_data <- rbind(processed_anc2 
                    %>% mutate(type = case_when(p282 == 0 & p297 == 0 & p300 == 0 ~ 'wt',
                                                p282 == 1 & p297 == 0 & p300 == 0 ~ '282',
                                                p282 == 0 & p297 == 1 & p300 == 0 ~ '297',
                                                p282 == 0 & p297 == 0 & p300 == 1 ~ '300',
                                                p282 == 1 & p297 == 1 & p300 == 0 ~ '282|297',
                                                p282 == 1 & p297 == 0 & p300 == 1 ~ '282|300',
                                                p282 == 0 & p297 == 1 & p300 == 1 ~ '297|300',
                                                p282 == 1 & p297 == 1 & p300 == 1 ~ '282|297|300',
                                                .default = 'ok')),
      processed_anc5 %>% mutate(type = "anc5"))

combo_data_processed <- combo_data %>% 
  select(-c(p282, p297, p300)) %>%
  mutate(type = fct_inorder(type),
         geno = paste0(p72, p193, p258, p271, p273),
         geno = fct_inorder(geno),
         norm_activity = log10(activity), # Log-transform
         norm_activity = norm_activity - mean(norm_activity[reference == 'L∆HFIALA' & type == 'wt'])
         # Normalize by the mean activity of the wt sequence
         ) %>% 
  ungroup() %>%
  select(-c("replicate", "activity")) %>%
  group_by(geno, type) %>%
  summarise(mean_activity = mean(norm_activity))
  
```

Then plot a similar heat map

```{r}
combo_data_processed %>%
  ggplot(aes(x = type, y = geno, fill = mean_activity)) +
  # Use 'color' for the border and 'linewidth' to make it stand out
  geom_tile(width = 0.9, height = 0.9, linewidth = 0.3) +
  # Blue-to-Red gradient
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", 
                       midpoint = 0) +
  theme_minimal() +
  theme(
    # Tilt x-axis labels 45 degrees
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    panel.grid = element_blank()
  ) +
  labs(fill = "Value", color = "Significant") +
  coord_fixed() # Keeps tiles square
```

Also figure grade layout

```{r}
combo_data_processed_for_plot <- combo_data_processed %>%
  mutate(type = case_when(type == "wt" ~ "AncDHCH2",
                          type == "282" ~ "+282",
                          type == "297" ~ "+297",
                          type == "300" ~ "+300",
                          type == "282|297" ~ "+282|297",
                          type == "282|300" ~ "+282|300",
                          type == "297|300" ~ "+297|300",
                          type == "282|297|300" ~ "+282|297|300",
                          type == "anc5" ~ "AncDHCH5")) %>%
  mutate(
    geno = fct_inorder(geno),
    type = fct_rev(fct_inorder(type)) # Reversed so AncDHCH2 is on top if desired
  )

tile <- combo_data_processed_for_plot %>%
  ggplot(aes(x = geno, y = type, fill = mean_activity)) +
  # Black border for significance, light grey for others to match the screenshot "grid" look
  geom_tile(color = "black", linewidth = 0.1) +
  # Moving X axis to the top
  scale_x_discrete(position = "top") +
  scale_fill_gradient2(
    low = "#1F78B4", mid = "white", high = "#E31A1C", 
    midpoint = 0
  ) +
  theme_minimal() +
  theme(
    # 45 degree tilt, adjusted for top position
    axis.text.x.top = element_text(angle = 45, hjust = 0, vjust = 0, size = 8),
    axis.title.y = element_blank(),
    axis.text.y = element_text(size = 8),
    axis.title.x.top = element_text(size = 8),
    panel.grid = element_blank(),
    legend.position = "right",
    legend.text = element_text(size = 8)
  ) +
  labs(x = "Genotype", 
       fill = expression("log"[10]*" fold change in activity")) +
  coord_fixed()

tile

# Clean figure
ggsave("fig_4_a.svg", plot = tile + theme(legend.position = "none"), width = 180/1.2, height = 247/3, dpi = 600, units = "mm")
# With legend for reference
ggsave("fig_4_a_legend.svg", plot = tile, width = 180/1.2, height = 247/3, dpi = 600, units = "mm")

write_csv(combo_data_processed_for_plot, "all_log10_fold_changes.csv")
```

Before moving to individual correlations I think a subset figure is useful for the manuscript

```{r}
tile <- combo_data_processed %>%
  filter(type %in% c("wt", "anc5")) %>%
  mutate(type = case_when(type == "wt" ~ "AncDHCH2",
                          type == "anc5" ~ "AncDHCH5")) %>%
  mutate(
    geno = fct_inorder(geno),
    type = fct_rev(fct_inorder(type)) # Reversed so AncDHCH2 is on top if desired
  ) %>%
  ggplot(aes(x = geno, y = type, fill = mean_activity)) +
  # Black border for significance, light grey for others to match the screenshot "grid" look
  geom_tile(color = "black", linewidth = 0.1) +
  # Moving X axis to the top
  scale_x_discrete(position = "top") +
  scale_fill_gradient2(
    low = "#1F78B4", mid = "white", high = "#E31A1C", 
    midpoint = 0
  ) +
  theme_minimal() +
  theme(
    # 45 degree tilt, adjusted for top position
    axis.text.x.top = element_text(angle = 45, hjust = 0, vjust = 0, size = 8),
    axis.title.y = element_blank(),
    axis.text.y = element_text(size = 8),
    axis.title.x.top = element_text(size = 8),
    panel.grid = element_blank(),
    legend.position = "right",
    legend.text = element_text(size = 8)
  ) +
  labs(x = "Genotype", 
       fill = expression("log"[10]*" fold change in activity")) +
  coord_fixed()



tile

# Clean figure
ggsave("fig_2_a.svg", plot = tile + theme(legend.position = "none"), width = 180/1.2, height = 247/4, dpi = 600, units = "mm")
# With legend for reference
ggsave("fig_2_a_legend.svg", plot = tile, width = 180/1.2, height = 247/4, dpi = 600, units = "mm")

```

and epistatic effects

```{r}
tile_2 <- ees_combined %>%
  filter(type %in% c("wt", "anc5")) %>%
  mutate(type = case_when(type == "wt" ~ "AncDHCH2",
                          type == "anc5" ~ "AncDHCH5")) %>%
  mutate(
    sig = ifelse(is.na(sig), FALSE, sig),
    geno = fct_inorder(geno),
    type = fct_rev(fct_inorder(type)) # Reversed so AncDHCH2 is on top if desired
  ) %>%
  ggplot(aes(x = geno, y = type, fill = new_val)) +
  # Black border for significance, light grey for others to match the screenshot "grid" look
  geom_tile(aes(color = sig), width = 0.9, height = 0.9, linewidth = 0.3) +
  # Moving X axis to the top
  scale_x_discrete(position = "top") +
  scale_fill_gradient2(
    low = "#1F78B4", mid = "white", high = "#E31A1C",
    midpoint = mean(ees_combined$new_val, na.rm = TRUE)
  ) +
  scale_color_manual(values = c("TRUE" = "black", "FALSE" = "grey90")) +
  theme_minimal() +
  theme(
    # 45 degree tilt, adjusted for top position
    axis.text.x.top = element_text(angle = 45, hjust = 0, vjust = 0, size = 8),
    axis.title.y = element_blank(),
    axis.text.y = element_text(size = 8),
    axis.title.x.top = element_text(size = 8),
    panel.grid = element_blank(),
    legend.position = "right",
    legend.text = element_text(size = 8)
  ) +
  labs(x = "Genotype", 
       fill = "Epistatic effect (EE)") +
  coord_fixed()

tile_2

# Clean figure
ggsave("fig_2_d.svg", plot = tile_2 + theme(legend.position = "none"), width = 180/1.2, height = 247/4, dpi = 600, units = "mm")
# With legend for reference
ggsave("fig_2_d_legend.svg", plot = tile_2, width = 180/1.2, height = 247/4, dpi = 600, units = "mm")

```

## Correlation to Anc5

Correlate wt with each, then anc5 with each

```{r}
combo_data_processed %>% 
  filter(type == 'wt' | type == '282') %>%
  pivot_wider(names_from = type, values_from = mean_activity) %>%
  ggplot(aes(x = wt, y = `282`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  xlim(-2,4) +
  ylim(-2,4) +
  stat_cor(label.y = 4, aes(label = after_stat(rr.label))) +
  theme_classic()
```

```{r}
combo_data_processed %>% 
  filter(type == 'wt' | type == '297') %>%
  pivot_wider(names_from = type, values_from = mean_activity) %>%
  ggplot(aes(x = wt, y = `297`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  xlim(-2,4) +
  ylim(-2,4) +
  stat_cor(label.y = 4, aes(label = after_stat(rr.label))) +
  theme_classic()
```

```{r}
combo_data_processed %>% 
  filter(type == 'wt' | type == '300') %>%
  pivot_wider(names_from = type, values_from = mean_activity) %>%
  ggplot(aes(x = wt, y = `300`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  xlim(-2,4) +
  ylim(-2,4) +
  stat_cor(label.y = 4, aes(label = after_stat(rr.label))) +
  theme_classic()
```

```{r}
combo_data_processed %>% 
  filter(type == 'wt' | type == '282|297') %>%
  pivot_wider(names_from = type, values_from = mean_activity) %>%
  ggplot(aes(x = wt, y = `282|297`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  xlim(-2,4) +
  ylim(-2,4) +
  stat_cor(label.y = 4, aes(label = after_stat(rr.label))) +
  theme_classic()
```

```{r}
combo_data_processed %>% 
  filter(type == 'wt' | type == '282|300') %>%
  pivot_wider(names_from = type, values_from = mean_activity) %>%
  ggplot(aes(x = wt, y = `282|300`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  xlim(-2,4) +
  ylim(-2,4) +
  stat_cor(label.y = 4, aes(label = after_stat(rr.label))) +
  theme_classic()
```

```{r}
combo_data_processed %>% 
  filter(type == 'wt' | type == '297|300') %>%
  pivot_wider(names_from = type, values_from = mean_activity) %>%
  ggplot(aes(x = wt, y = `297|300`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  xlim(-2,4) +
  ylim(-2,4) +
  stat_cor(label.y = 4, aes(label = after_stat(rr.label))) +
  theme_classic()
```

```{r}
combo_data_processed %>% 
  filter(type == 'wt' | type == '282|297|300') %>%
  pivot_wider(names_from = type, values_from = mean_activity) %>%
  ggplot(aes(x = wt, y = `282|297|300`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  xlim(-2,4) +
  ylim(-2,4) +
  stat_cor(label.y = 4, aes(label = after_stat(rr.label))) +
  theme_classic()
```

```{r}
combo_data_processed %>% 
  filter(type == 'wt' | type == 'anc5') %>%
  pivot_wider(names_from = type, values_from = mean_activity) %>%
  ggplot(aes(x = `wt`, y = `anc5`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  xlim(-2,4) +
  ylim(-2,4) +
  stat_cor(label.y = 4, aes(label = after_stat(rr.label))) +
  theme_classic()
```

```{r}
combo_data_processed %>% 
  filter(type == 'anc5' | type == '282') %>%
  pivot_wider(names_from = type, values_from = mean_activity) %>%
  ggplot(aes(x = `anc5`, y = `282`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  xlim(-2,4) +
  ylim(-2,4) +
  stat_cor(label.y = 4, aes(label = after_stat(rr.label))) +
  theme_classic()
```

```{r}
combo_data_processed %>% 
  filter(type == 'anc5' | type == '297') %>%
  pivot_wider(names_from = type, values_from = mean_activity) %>%
  ggplot(aes(x = `anc5`, y = `297`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  xlim(-2,4) +
  ylim(-2,4) +
  stat_cor(label.y = 4, aes(label = after_stat(rr.label))) +
  theme_classic()
```

```{r}
combo_data_processed %>% 
  filter(type == 'anc5' | type == '300') %>%
  pivot_wider(names_from = type, values_from = mean_activity) %>%
  ggplot(aes(x = `anc5`, y = `300`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  xlim(-2,4) +
  ylim(-2,4) +
  stat_cor(label.y = 4, aes(label = after_stat(rr.label))) +
  theme_classic()
```

```{r}
combo_data_processed %>% 
  filter(type == 'anc5' | type == '282|297') %>%
  pivot_wider(names_from = type, values_from = mean_activity) %>%
  ggplot(aes(x = `anc5`, y = `282|297`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  xlim(-2,4) +
  ylim(-2,4) +
  stat_cor(label.y = 4, aes(label = after_stat(rr.label))) +
  theme_classic()
```

```{r}
two_enough <- combo_data_processed %>% 
  filter(type == 'anc5' | type == '282|300') %>%
  pivot_wider(names_from = type, values_from = mean_activity) %>%
  ggplot(aes(x = `282|300`, y = `anc5`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "#E31A1C") +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  xlim(-2,2) +
  ylim(-2,2) +
  stat_cor(size = 2.8, label.y = 2, aes(label = after_stat(rr.label))) +
  theme_classic() +
  labs(y = expression("AncDHCH5 ∆log"[10]*" activity"),
       x = expression("+282|300 ∆log"[10]*" activity")) +
  theme(
    axis.text.x = element_text(size = 8),
    axis.text.y = element_text(size = 8),
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    panel.grid.minor = element_blank()
  )

two_enough

ggsave("282|300_vs_anc5.svg", plot = two_enough, width = 180/4, height = 247/4.4, dpi = 600, units = "mm")

```

```{r}
combo_data_processed %>% 
  filter(type == 'anc5' | type == '297|300') %>%
  pivot_wider(names_from = type, values_from = mean_activity) %>%
  ggplot(aes(x = `anc5`, y = `297|300`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  xlim(-2,4) +
  ylim(-2,4) +
  stat_cor(label.y = 4, aes(label = after_stat(rr.label))) +
  theme_classic()
```

```{r}
combo_data_processed %>% 
  filter(type == 'anc5' | type == '282|297|300') %>%
  pivot_wider(names_from = type, values_from = mean_activity) %>%
  ggplot(aes(x = `anc5`, y = `282|297|300`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  xlim(-2,4) +
  ylim(-2,4) +
  stat_cor(label.y = 4, aes(label = after_stat(rr.label))) +
  theme_classic()
```

Trying to align epistasis coefficients as well

```{r}
ees_combined %>%
  filter(sig & (type == 'wt' | type == '282')) %>%
  pivot_wider(names_from = type, values_from = new_val) %>%
  ggplot(aes(x = wt, y = `282`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  xlim(-3,4) +
  ylim(-3,4) +
  stat_cor(label.y = 4, aes(label = after_stat(rr.label))) +
  theme_classic()
```

```{r}
ees_combined %>%
  filter(sig & (type == 'wt' | type == '297')) %>%
  pivot_wider(names_from = type, values_from = new_val) %>%
  ggplot(aes(x = wt, y = `297`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  xlim(-3,4) +
  ylim(-3,4) +
  stat_cor(label.y = 4, aes(label = after_stat(rr.label))) +
  theme_classic()
```

```{r}
ees_combined %>%
  filter(sig & (type == 'wt' | type == '300')) %>%
  pivot_wider(names_from = type, values_from = new_val) %>%
  ggplot(aes(x = wt, y = `300`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  xlim(-3,4) +
  ylim(-3,4) +
  stat_cor(label.y = 4, aes(label = after_stat(rr.label))) +
  theme_classic()
```

```{r}
ees_combined %>%
  filter(sig & (type == 'wt' | type == '282|297')) %>%
  pivot_wider(names_from = type, values_from = new_val) %>%
  ggplot(aes(x = wt, y = `282|297`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  xlim(-3,4) +
  ylim(-3,4) +
  stat_cor(label.y = 4, aes(label = after_stat(rr.label))) +
  theme_classic()
```

```{r}
ees_combined %>%
  filter(sig & (type == 'wt' | type == '282|300')) %>%
  pivot_wider(names_from = type, values_from = new_val) %>%
  ggplot(aes(x = wt, y = `282|300`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  xlim(-3,4) +
  ylim(-3,4) +
  stat_cor(label.y = 4, aes(label = after_stat(rr.label))) +
  theme_classic()
```

```{r}
ees_combined %>%
  filter(sig & (type == 'wt' | type == '297|300')) %>%
  pivot_wider(names_from = type, values_from = new_val) %>%
  ggplot(aes(x = wt, y = `297|300`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  xlim(-3,4) +
  ylim(-3,4) +
  stat_cor(label.y = 4, aes(label = after_stat(rr.label))) +
  theme_classic()
```

```{r}
ees_combined %>%
  filter(sig & (type == 'wt' | type == '282|297|300')) %>%
  pivot_wider(names_from = type, values_from = new_val) %>%
  ggplot(aes(x = wt, y = `282|297|300`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  xlim(-3,4) +
  ylim(-3,4) +
  stat_cor(label.y = 4, aes(label = after_stat(rr.label))) +
  theme_classic()
```

```{r}
ees_combined %>%
  select(c(geno, type, new_val, sig)) %>%
  filter(sig & (type == 'anc5' | type == '282|297|300')) %>%
  pivot_wider(names_from = type, values_from = new_val) %>%
  ggplot(aes(x = anc5, y = `282|297|300`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  xlim(-3,4) +
  ylim(-3,4) +
  stat_cor(label.y = 4, aes(label = after_stat(rr.label))) +
  theme_classic()
```

```{r}
ees_combined %>%
  select(c(geno, type, new_val, sig)) %>%
  filter(sig & (type == 'wt' | type == 'anc5')) %>%
  pivot_wider(names_from = type, values_from = new_val) %>%
  ggplot(aes(x = wt, y = anc5)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  xlim(-3,4) +
  ylim(-3,4) +
  stat_cor(label.y = 4, aes(label = after_stat(rr.label))) +
  theme_classic()
```
